Acceleration of Derivative Calculations with Application to Radial Basis Function -- Finite-Differences on the MIC Architecture

### 2013-10-19 ieee_paper.tex
changes.

\begin{document}
\title{Sparse Matrix Vector Multiplication with Multiple vectors and
  Multiple Matrices on the MIC Architecture \todo{? more application oriented''}}

\author{\IEEEauthorblockN{Gordon Erlebacher\IEEEauthorrefmark{1},
Erik Saule\IEEEauthorrefmark{2}, Natasha Flyer\IEEEauthorrefmark{3}, 
and Evan Bollig\IEEEauthorrefmark{1}}
\IEEEauthorblockA{\IEEEauthorrefmark{1}Department of Scientific Computing, 
Florida State University}
\IEEEauthorblockA{\IEEEauthorrefmark{2}Department of Computer Science, University of North Carolina at Charlotte}
\IEEEauthorblockA{\IEEEauthorrefmark{3}Computational and Information Systems Laboratory, UCAR}
Email: gerlebacher@fsu.edu, esaule@uncc.edu, flyer@ucar.edu, bollig@gmail.com}
\maketitle

@@

In this work, we use an Intel Xeon Phi 5110P coprocessor. This card
has 8 memory controllers that can each execute 5 billion
transactions per second. Each has two 32-bit channels, achieving a total
bandwidth of 320~GB/s aggregated across all the memory
controllers. There are 60 cores clocked at 1.05~GHz. Their memory
interfaces are 32-bit wide with two channels and the total bandwidth
is 8.4~GB/s per core. Thus, the cores should be able to consume 504~GB/s
at most. However, the bandwidth between the cores and the memory
controllers is limited by the ring network that connects the cores and
the memory controller. Its precise bandwidth is unknown but it is
believed to be between 200~GB/s and 250~GB/s.

@@

As explained in Section~\ref{sec:model}, we expect to achieve a performance of
about 210~Gflop/s in single precision in the best case. We will focus
on the single precision case with four vectors and four matrices. 
Similar techniques apply for other matrix/vector combinations. This target
performance represents 10\% of the peak performance of the
architecture which is impossible to reach without an efficient vectorization of
the kernel. Indeed, to reach such a performance, a Fused Multiply-Add
instruction on fully loaded registers must be executed at most every
10 cycles.

In term of memory layout, we adopt a format similar to the ELLPACK
format to store the matrix since the number of nonzeros per row is
constant throughout the matrix. The matrix is given in two arrays. The
first one is the classical {\tt col\_id} array that describes the
sparsity structure of the matrix. Each row of {\tt col\_id} consists
of $n_z$ entries that label the column index of each nonzero in
$A$. The other array is the {\tt data} array that gives the values of
the nonzeros in the four matrices. The values of the matrices are
interleaved by group of four so that the value $A^l_{col\_id[k]}$ of
the $k$th nonzero of the $l$th matrix is at index $\floor{\frac{k-1}{4}}*4*n_m + (l-1)*4+(k-1)\%4$.
(In other words, the first 4 entries are from $A^{1}$ and the next 4 are from $A^2$.) The vectors {\tt x} are
interleaved so that {\tt x}$^l_i$ and {\tt x}$^{l+1}_i$ are
consecutive in memory. The vector {\tt y} is similarly interleaved,
grouped in sets of 16, which correspond to four derivatives applied to
four functions, at a single point.

@@

The core of the technique is the use of swizzling and permutation
features of the vectors in the Intel MIC architecture. A vector of 512
bits is composed of four lanes (sometimes called channels) of 128
bits, which are made of four segments of 32 bits. A swizzling
operation allows the reorder of the elements within each lane (see how
{\tt mat} is extracted from {\tt mat\_ent} in
Figure~\ref{code:mat_mul}). While a permutation reorders entire lanes,
without changing their contents (see for instance how {\tt vect} is
generated from {\tt all\_vect} in Figure~\ref{code:mat_mul}).  Both
permutation and swizzling support common reordering and broadcast. In
the code, the ``permute'' function needs to use the {\tt
  \_mm512\_permute4f128\_epi32} instruction which is only defined on
integers and therefore in place typecasts are necessary to convert to
a vector to/from the integer type.  (Note that we prefer swizzling to
permutation if possible because we believe it is cheaper. Many
instructions have swizzling capability, while permuation is its own
instruction. \todo{do we need a reference to intel Xeon phi instruction set?})

>>>>

\begin{document}
\title{Acceleration of Derivative Calculations with Application to Radial Basis Function --
Finite-Differences on the MIC Architecture}
%\title{Sparse Matrix Vector Multiplication with Multiple vectors and
  %Multiple Matrices on the MIC Architecture \todo{? more application oriented''}}

\author{\IEEEauthorblockN{Gordon Erlebacher\IEEEauthorrefmark{1},
Erik Saule\IEEEauthorrefmark{2}, Natasha Flyer\IEEEauthorrefmark{3}, 
and Evan Bollig\IEEEauthorrefmark{1}}
\IEEEauthorblockA{\IEEEauthorrefmark{1}Department of Scientific Computing, 
Florida State University\\ %, Tallahassee %FL 32306-4120\\
Email: gerlebacher@fsu.edu, bollig@gmail.com}
\IEEEauthorblockA{\IEEEauthorrefmark{2}Department of Computer Science, University of North Carolina at Charlotte\\
Email: erik.saule@uncc.edu}
\IEEEauthorblockA{\IEEEauthorrefmark{3}Computational and Information Systems Laboratory, UCAR \\
Email: flyer@ucar.edu}}
%\IEEEauthorblockA{\IEEEauthorrefmark{4}Department of Scientific Computing, Florida State University\\
%Email: bollig@gmail.com}\todo{check affiliation fsu's are strange.}}
\maketitle

@@

In this work, we use an Intel Xeon Phi 5110P coprocessor. This card
has 8 memory controllers that can each execute 5 billion
transactions per second. Each has two 32-bit channels, achieving a total
bandwidth of 320~GB/s aggregated across all the memory
controllers. There are 60 cores clocked at 1.05~GHz. Their memory
interfaces are 32-bit wide with two channels and the total bandwidth
is 8.4~GB/s per core. Thus, the cores should be able to consume 504~GB/s
at most. However, the bandwidth between the cores and the memory
controllers is limited by the ring network that connect the cores and
the memory controller. Its precise bandwidth is unknown but it is
believed to be between 200~GB/s and 250~GB/s.

@@

As explained in Section~\ref{sec:model}, we expect to achieve a performance of
about 210~Gflop/s in single precision in the best case. We will focus
on the single precision case with four vectors and four matrices. 
Similar techniques apply for other matrix/vector combinations. This target
performance represents 10\% of the peak performance of the
architecture which is impossible without an efficient vectorization of
the kernel. Indeed, to reach such a performance, a Fused Multiply-Add
instruction on fully loaded registers must be executed at most every
10 cycles.

In term of memory layout, we adopt a format similar to the ELLPACK
format to store the matrix since the number of nonzeros per row is
constant throughout the matrix. The matrix is given in two arrays. The
first one is the classical {\tt col\_id} array that describes the
sparsity structure of the matrix. Each row of {\tt col\_id} consists
of $n_z$ entries that label the column index of each nonzero in
$A$. The other array is the {\tt data} array that gives the values of
the nonzeros in the four matrices. The values of the matrices are
interleaved by group of four so that the value $A^l_{col\_id[k]}$ of
the $k$th nonzero of the $l$th matrix is at index $\floor{\frac{k-1}{4}}*4*n_m + (l-1)*4+(k-1)\%4$.\todo{check that formula}
(In other words, the first 4 entries are from $A^{1}$ and the next 4 are from $A^2$.) The vectors {\tt x} are
interleaved so that {\tt x}$^l_i$ and {\tt x}$^{l+1}_i$ are
consecutive in memory. The vector {\tt y} is similarly interleaved,
grouped in sets of 16, which correspond to four derivatives applied to
four functions, at a single point.

@@

The core of the technique is the use of swizzling and permutation
features of the vectors in the Intel MIC architecture. A vector of 512
bits is composed of four lanes (sometimes called channels) of 128
bits, which are made of four segments of 32 bits. A swizzling
operation allows the reorder of the elements within each lane (see how
{\tt mat} is extracted from {\tt mat\_ent} in
Figure~\ref{code:mat_mul}). While a permutation reorders entire lanes,
without changing their contents (see for instance how {\tt vect} is
generated from {\tt all\_vect} in Figure~\ref{code:mat_mul}).  Both
permutation and swizzling support common reordering and broadcast. In
the code, the ``permute'' function needs to use the {\tt
  \_mm512\_permute4f128\_epi32} instruction which is only defined on
integers and therefore in place typecasts are necessary to convert to
a vector to/from the integer type.  (Note that we prefer swizzling to
permutation if possible because we believe it is cheaper. Many
instructions have swizzling capability, while permuation is its own
instruction. %\todo{do we need a reference to intel Xeon phi instruction set?})
